SAT Solving for Vesicle Traffic Systems in Cells

### 2017-07-20 paper/encoding.tex
merged

We translate the search problem into a Boolean satisfiability
problem and use SAT solvers to find the satisfying VTSs.
%
We will first present the variables used to encode the
VTSs and the properties.
%
Afterwards, we will present the formulas corresponding to the
properties.

@@

\item Boolean variable $n_{i,m}$ indicates if $m \in \nlabel(i)$
\item Boolean variable $e_{i,j,q}$ indicates if $q$th edge exists between $i$ and $j$.
\item Boolean variable $e_{i,j,q,m}$ indicates if $q$th edge between $i$ and $j$ contains molecule $m$.

\item Boolean variable $p_{k,k'}$ indicates if $(m,m') \in \pairs$
\item uninterpreted Boolean functions $\nodef_m : \booleans^\mu \maps \booleans$
encoding $\nodef(m)$ map
\item uninterpreted Boolean functions $\edgef_m : \booleans^\mu \maps \booleans$
encoding $\edgef(m)$ map
\end{enumerate}
We also have auxiliary Boolean variables that will help us encode the well-fused and stability properties 
\begin{enumerate}
\item $a_{i,m}$ indicates that molecule $m$ is active at node $i$, i.e., $\nodef(m,L(i))$
  holds
\item $b_{i,j,q,m}$ indicates that molecule $m$ is active at $q$th edge $(i,M',j)$ between $i$ and $j$, i.e., $\edgef(m,M')$ holds
\item $r_{i,j,m,p}$ indicates that there is a $m$-path from
  $i$ to $j$ of length less than or equal to $p$
\end{enumerate}
For $k$-connected property, we also use the following auxiliary Boolean variables
\begin{enumerate}
\item $d_{i,j,q}$ indicates $q$th edge between $i$ and $j$ is dropped
\item $r'_{i,j}$ indicates that there is a path from $i$ to $j$ in the modified VTS
\end{enumerate}

We will describe the Boolean formulas that encode VTSs in several categories.
%
In the end, we will present in a table the formulas needed for the
model variants.
%
To avoid cumbersome notation, we will not explicitly write the ranges of the indexing
in the formulas.
%
$i$ and $j$ will be ranged over nodes.
%
$m$ will be ranged over molecules.
%
$q$ will be ranged over edges between two nodes.
%

@@

\subsubsection{Well-fused constraints}
Boolean constraint~\eqref{eq:fuse1} encodes that each edge must fuse with
its destination node.
%
Boolean constraint~\eqref{eq:fuse2} encodes that each edge should not
be able to potentially fuse with any node other than its destination node.
\begin{align}
  \bigwedge\limits_{i,j,q} e_{i,j,q} \limplies \bigvee_{m,m^{\prime}} (b_{i,j,q,m} \land a_{j,m^{\prime}} \land p_{m,m^{\prime}})
  \tag{V7}\label{eq:fuse1}  \\
\bigwedge\limits_{i,j,q,m} b_{i,j,q,m} \limplies \neg \bigvee_{j \neq j^{\prime}, m^{\prime\prime}} ( a_{j^{\prime},m^{\prime\prime}} \land p_{m,m^{\prime\prime}})
  \tag{V8}\label{eq:fuse2}  
\end{align}

@@

\subsubsection{Regulation on nodes and edges}
We are considering several models that differ in constraints on
the activity of molecules.
%
We will present~\eqref{eq:ann}-~\eqref{eq:aep} that encodes
the varying constraints.
%
The following formula encodes no conditions on activities on nodes,
i.e., all the present molecules on the nodes are active.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} = a_{i,m}    \tag{Ann}\label{eq:ann}
\end{align}
The following formula encodes that activity of a molecule $m$ on the node is
defined by a Boolean function $\nodef_m$ of presence of molecules present on that node.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} \limplies a_{i,m} =  \nodef_m (n_{i,1},\dots,n_{i,\mu}) 
\tag{Anb}\label{eq:anb}
\end{align}
The following formula encodes no conditions on activities on edges,
i.e., all the present molecules on the edges are active.
\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} = b_{i,j,q,m}
\tag{Aen}\label{eq:aen}
\end{align}
The following formula encodes that activity of a molecule $m$ on the edge is
defined by a Boolean function $\edgef_m$ of presence of molecules present on that edge.
\begin{align}
   \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies b_{i,j,q,m} = \edgef_k(e_{i,j,q,1}, .., e_{i,j,q,\mu} )
  \tag{Aeb}\label{eq:aeb}
\end{align}
%
The following formula encodes that the activity of the molecules on
edges is defined by inhibition by other molecules based on the pairing
matrix. 
\begin{align}
   \bigwedge\limits_{i,j,q, m}  [e_{i,j,q,m} \limplies  \bigwedge_{m' \neq m} (p_{m,m'} \limplies e_{i,j,q,m'})] \lequiv (\neg b_{i,j,q,m} \land\;\;\;\;  \bigwedge_{\mathclap{m' \neq m, p_{m,m'}}} \neg b_{i,j,q,m'})
  \tag{Aep}\label{eq:aep}
\end{align}
%
\subsubsection{Constraints for stability condition}
%
We use $m$-reachability to encode the stability condition in VTSs.
%
The following formula recursively encodes that node $j$ is $m$-reachable from node $i$ in less than $p$ steps
if either there is a direct edge between $i$ and $j$ with $m$ present on the edge or there is a edge between $i^{\prime \prime}$ and
$(i \neq i^{\prime \prime})$ with $m$ present, and j is $m$-reachable from $i'$ in less than $p-1$ steps.
%
\begin{align}
  \bigwedge\limits_{i,j,m,p} r_{i,j,m,p} \limplies (\bigvee_{q} \, e_{i,j,q,m} \lor \bigvee_{i\neq i^{\prime}} ( \, \bigvee_{q} e_{i,i^{\prime},q,m}) \land r_{i^{\prime},j,m,p-1} )
  \tag{R1}\label{eq:reach1}
\end{align}
Now we can encode stability using the reachability variables.
and say if there is an $m$-edge between $i$ and $j$, there is
$m$-reachable path from $j$ to $i$.
\begin{align}
 \bigwedge\limits_{i,j,m} (\bigvee_{q} e_{i,j,q,m}) \limplies r_{j,i,m,|N|}
  \tag{R2}\label{eq:reach2}
\end{align}

@@

\subsubsection{$k$-connectivity constraints}
To check whether $k$-connected is a necessary condition, we remove (drop) $k-1$ edges from the graph and if it
disconnects the graph, and we get an assignment. We have a graph that is not  $k$-connected.
%
Boolean constraints~\ref{eq:drop1}-\ref{eq:drop4}
encode the relevant constraints for reachability
in the modified VTS. 
%
The following formulas encode that only
existing edges can be dropped and exactly $k-1$ edges are dropped.
\begin{align}
  \bigwedge\limits_{i,j,q} d_{i,j,q} \limplies e_{i,j,q}  \tag{D1}\label{eq:drop1}\\
  \sum_{i,j,q} d_{i,j,q} = k-1
  \tag{D2}\label{eq:drop2}
\end{align}
We need to define reachability in the modified VTS, therefore we use
a new variable $r'_{i,j}$ to encode reachability from $i$ to $j$.
In the following formula, we encode $r'_{i,j}$ is true if there is an
edge $(i,\_,j)$ and it is not dropped, or there is a node
$i^{\prime}$ such that, there is an edge $(i,\_,i^{\prime})$ which is
not dropped and $r'_{i',j}$ is true.
\begin{align}
\bigwedge\limits_{i,j}  [\bigvee_{q} (e_{i,j,q} \land  \neg d_{i,j,q}) \lor  (\bigvee_{i' \neq i}  r^{\prime}_{i',j} \land  \bigvee_{q} (e_{i,i',q} \land \neg d_{i,i',q}) ] \limplies r^{\prime}_{i,j}  
  \tag{D3}\label{eq:drop3}
\end{align}
Since we are looking for a graph that is not $k$-connected,
the following formula encodes that there are nodes $i$ and $j$ such that
there is no path between them in the underlying undirected graph.
\begin{align}
   \bigvee\limits_{i,j} \neg (r^{\prime}_{i,j} \lor r^{\prime}_{j,i})
  \tag{D4}\label{eq:drop4}
\end{align}

The key improvement in our work over the earlier tool is the
encoding of reachability, which was done using enumeration of paths.
%
In the current work, we have encoded reachability in two different
ways in formulas~\eqref{eq:reach1} and~\eqref{eq:drop3}.
%
The reachability is recursively defined in~\eqref{eq:drop3} and has
trivial solutions by making all $r'$s true.
%
However, the trivial solutions are disallowed by formula~\eqref{eq:drop4} and we find
only the solutions that captures the evidence of unreachability.
%
On the other hand,
we have added length of paths in our reachability encoding in formula~\eqref{eq:reach1},
which needs relatively more auxiliary variables.
%
This is because formula~\eqref{eq:reach2} has only positive
occurrences of the reachability variables and if we had defined
$r$s in ~\eqref{eq:reach1} without paths,
the circular dependencies in the recursive definitions of $r$s
may have resulted in spurious satisfying assignments that
do not encode reachability.
%
By adding the path length, we break the circular dependencies, the
formula remains polynomial in size, and satisfying assignments only
correspond to correct reachability.

@@

% We have applied the tool for the above variations for searching the
% VTSs that satisfies their corresponding properties. 
%
% After creating the formula using different Boolean variables 
% we ship them directly to the solver.
% %
% In case of a satisfying (SAT) assignment of the added constraints the
% Z3 theorem prover gives a model with the variable assignment.
% %
% In case no possible satisfying assignment the solver returns UNSAT.

@@

% \subsection{Creating a solver and solving these constraints}
% We will create a solver and add these build formula using these constraints to that and then check whether there a satisfiable assignment that satisfies all these constraints.

% \begin{figure}[ht]
% %\vspace{-3mm}
% \begin{verbatim}
% # Create a solver
% s = Solver()
% # Add these formulas to it.
% s.add(A0, A1, C0, C1, C2, C4, C5, F0, F1, F2, F3, F4, F5, D0, D1, D2, D3)

>>>>

We translate the search problem into a Boolean satisfiability
problem and use SAT solvers to find the satisfying VTSs.
%
We will first present the variables used to encode the
VTSs and the properties.
%
Afterwards, we will present the constraints corresponding to the
properties.

@@

\item Boolean variable $n_{i,m}$ indicates if $m \in \nlabel(i)$
\item Boolean variable $e_{i,j,q}$ indicates if $q$th edge exists between $i$ and $j$.
\item Boolean variable $e_{i,j,q,m}$ indicates if $q$th edge between $i$ and $j$ contains $m$.
\item Boolean variable $p_{m,m'}$ indicates if $(m,m') \in \pairs$
\item uninterpreted Boolean functions $\nodef_m : \booleans^\mu \maps \booleans$
encoding $\nodef(m)$ map
\item uninterpreted Boolean functions $\edgef_m : \booleans^\mu \maps \booleans$
encoding $\edgef(m)$ map
\end{enumerate}
We also have auxiliary Boolean variables that will help us encode the well-fused and stability properties 
\begin{enumerate}
\item $a_{i,m}$ indicates that molecule $m$ is active at node $i$, i.e., $\nodef(m,L(i))$
  holds
\item $b_{i,j,q,m}$ indicates that molecule $m$ is active at $q$th edge $(i,M',j)$ between $i$ and $j$, i.e., $\edgef(m,M')$ holds
\item $r_{i,j,m,p}$ indicates if there is an $m$-path from
  $i$ to $j$ of length less than or equal to $p$
\end{enumerate}
For $k$-connected property, we also use the following auxiliary Boolean variables
\begin{enumerate}
\item $d_{i,j,q}$ indicates $q$th edge between $i$ and $j$ is dropped
\item $r'_{i,j}$ indicates if there is a path from $i$ to $j$ in the modified VTS
\end{enumerate}

We will describe the Boolean constraints that encode VTSs in several categories.
%
In the end, we will present in a table the constraints needed for the
model variants.
%
To avoid cumbersome notation, we will not explicitly write the ranges of the indexing
in the constraints.
%
$i$ and $j$ will range over nodes.
%
$m$ will range over molecules.
%
$q$ will range over edges between two nodes.
%

@@

\subsubsection{Well-fused constraints}
Constraint~\eqref{eq:fuse1} encodes that each edge must fuse with
its destination node.
%
Constraint~\eqref{eq:fuse2} encodes that each edge should not
be able to potentially fuse with any node other than its destination node.
\begin{align}
  \bigwedge\limits_{i,j,q} e_{i,j,q} \limplies \bigvee_{m,m^{\prime}} (b_{i,j,q,m} \land a_{j,m^{\prime}} \land p_{m,m^{\prime}})
  \tag{V7}\label{eq:fuse1}  \\
\bigwedge\limits_{i,j,q,m} b_{i,j,q,m} \limplies \neg \bigvee_{j \neq j^{\prime}, m^{\prime\prime}} ( a_{j^{\prime},m^{\prime\prime}} \land p_{m,m^{\prime\prime}})
  \tag{V8}\label{eq:fuse2}  
\end{align}

@@

\subsubsection{Regulation on nodes and edges}
We are considering several models that differ in constraints on
the activity of molecules.
%
We will present~\eqref{eq:ann}-~\eqref{eq:aep} that encodes
the varying constraints.
%
The following constraint encodes no conditions on activities on nodes,
i.e., all the present molecules on the nodes are active.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} = a_{i,m}    \tag{Ann}\label{eq:ann}
\end{align}
The following constraint encodes that activity of a molecule $m$ on the node is
defined by a Boolean function $\nodef_m$ of presence of molecules present on that node.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} \limplies a_{i,m} =  \nodef_m (n_{i,1},\dots,n_{i,\mu}) 
\tag{Anb}\label{eq:anb}
\end{align}
The following constraint encodes no conditions on activities on edges,
i.e., all the present molecules on the edges are active.
\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} = b_{i,j,q,m}
\tag{Aen}\label{eq:aen}
\end{align}
The following constraint encodes that activity of a molecule $m$ on the edge is
defined by a Boolean function $\edgef_m$ of presence of molecules present on that edge.
\begin{align}
   \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies b_{i,j,q,m} = \edgef_k(e_{i,j,q,1}, .., e_{i,j,q,\mu} )
  \tag{Aeb}\label{eq:aeb}
\end{align}
%
The following constraint encodes that the activity of the molecules on
edges is defined by inhibition by other molecules based on the pairing
matrix. 
\begin{align}
   \bigwedge\limits_{i,j,q, m}  [e_{i,j,q,m} \limplies  \bigwedge_{m' \neq m} (p_{m,m'} \limplies e_{i,j,q,m'})] \lequiv (\neg b_{i,j,q,m} \land\;\;\;\;  \bigwedge_{\mathclap{m' \neq m, p_{m,m'}}} \neg b_{i,j,q,m'})
  \tag{Aep}\label{eq:aep}
\end{align}
%
\subsubsection{Constraints for stability condition}
%
We use $m$-reachability to encode the stability condition in VTSs.
%
The following constraint recursively encodes that node $j$ is $m$-reachable from node $i$ in less than $p$ steps
if either there is a direct edge between $i$ and $j$ with $m$ present on the edge or there is a edge between $i^{\prime \prime}$ and
$(i \neq i^{\prime \prime})$ with $m$ present, and j is $m$-reachable from $i'$ in less than $p-1$ steps.
%
\begin{align}
  \bigwedge\limits_{i,j,m,p} r_{i,j,m,p} \limplies (\bigvee_{q} \, e_{i,j,q,m} \lor \bigvee_{i\neq i^{\prime}} ( \, \bigvee_{q} e_{i,i^{\prime},q,m}) \land r_{i^{\prime},j,m,p-1} )
  \tag{R1}\label{eq:reach1}
\end{align}
Now we can encode stability using the reachability variables.
and say if there is an $m$-edge between $i$ and $j$, there is
$m$-reachable path from $j$ to $i$.
\begin{align}
 \bigwedge\limits_{i,j,m} (\bigvee_{q} e_{i,j,q,m}) \limplies r_{j,i,m,|N|}
  \tag{R2}\label{eq:reach2}
\end{align}

@@

\subsubsection{$k$-connectivity constraints}
To check whether $k$-connected is a necessary condition, we remove (drop) $k-1$ edges from the graph and if it
disconnects the graph, and we get an assignment. We have a graph that is not  $k$-connected.
%
Boolean constraints~\ref{eq:drop1}-\ref{eq:drop4}
encode the relevant constraints for reachability
in the modified VTS. 
%
The following constraints encode that only
existing edges can be dropped and exactly $k-1$ edges are dropped.
\begin{align}
  \bigwedge\limits_{i,j,q} d_{i,j,q} \limplies e_{i,j,q}  \tag{D1}\label{eq:drop1}\\
  \sum_{i,j,q} d_{i,j,q} = k-1
  \tag{D2}\label{eq:drop2}
\end{align}
We need to define reachability in the modified VTS, therefore we use
a new variable $r'_{i,j}$ to encode reachability from $i$ to $j$.
In the following constraint, we encode $r'_{i,j}$ is true if there is an
edge $(i,\_,j)$ and it is not dropped, or there is a node
$i^{\prime}$ such that, there is an edge $(i,\_,i^{\prime})$ which is
not dropped and $r'_{i',j}$ is true.
\begin{align}
\bigwedge\limits_{i,j}  [\bigvee_{q} (e_{i,j,q} \land  \neg d_{i,j,q}) \lor  (\bigvee_{i' \neq i}  r^{\prime}_{i',j} \land  \bigvee_{q} (e_{i,i',q} \land \neg d_{i,i',q}) ] \limplies r^{\prime}_{i,j}  
  \tag{D3}\label{eq:drop3}
\end{align}
Since we are looking for a graph that is not $k$-connected,
the following constraint encodes that there are nodes $i$ and $j$ such that
there is no path between them in the underlying undirected graph.
\begin{align}
   \bigvee\limits_{i,j} \neg (r^{\prime}_{i,j} \lor r^{\prime}_{j,i})
  \tag{D4}\label{eq:drop4}
\end{align}

The key improvement in our work over the earlier tool is the
encoding of reachability, which was done using enumeration of paths.
%
In the current work, we have encoded reachability in two different
ways in constraints~\eqref{eq:reach1} and~\eqref{eq:drop3}.
%
The reachability is recursively defined in~\eqref{eq:drop3} and has
trivial solutions by making all $r'$s true.
%
However, the trivial solutions are disallowed by constraint~\eqref{eq:drop4} and we find
only the solutions that captures the evidence of unreachability.
%
On the other hand,
we have added length of paths in our reachability encoding in constraint~\eqref{eq:reach1},
which needs relatively more auxiliary variables.
%
This is because constraint~\eqref{eq:reach2} has only positive
occurrences of the reachability variables and if we had defined
$r$s in ~\eqref{eq:reach1} without paths,
the circular dependencies in the recursive definitions of $r$s
may have resulted in spurious satisfying assignments that
do not encode reachability.
%
By adding the path length, we break the circular dependencies, the
constraint remains polynomial in size, and satisfying assignments only
correspond to correct reachability.

@@

% We have applied the tool for the above variations for searching the
% VTSs that satisfies their corresponding properties. 
%
% After creating the constraint using different Boolean variables 
% we ship them directly to the solver.
% %
% In case of a satisfying (SAT) assignment of the added constraints the
% Z3 theorem prover gives a model with the variable assignment.
% %
% In case no possible satisfying assignment the solver returns UNSAT.

@@

% \subsection{Creating a solver and solving these constraints}
% We will create a solver and add these build constraint using these constraints to that and then check whether there a satisfiable assignment that satisfies all these constraints.

% \begin{figure}[ht]
% %\vspace{-3mm}
% \begin{verbatim}
% # Create a solver
% s = Solver()
% # Add these constraints to it.
% s.add(A0, A1, C0, C1, C2, C4, C5, F0, F1, F2, F3, F4, F5, D0, D1, D2, D3)
